Structured Genealogy Processes

### 2022-02-04 ms.Rnw
sgp updates

All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: no population event or simultaneous events, a birth from Deme $j$ to $j'$, a sample of Deme $j'$, and a migration from Deme $j$ to $j'$ where $j\neq j'$. Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample, which was in Deme $j''$. For $q>1$, we then have:

\begin{enumerate}[(a)]
  \item If $t_k\notin\halfopen{a_q,s_q}$, or if $u_k\notin\Bset\cup\Sset\cup\Mset$, so we have $\phi_{qk}^{j''}=1$.
  \item If $t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1})$, then again, $\phi_{qk}^{j''}=1$.
  \item If $j''\notin\{j,j'\}$, then the new lineage sample was not involved in any population events involving Demes $j$ and $j'$ at time $t_k$ and $\phi_{qk}^{j''}=1$.
  \item \label{item:M2Z} If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Mset^{j,j'}$, then there was a migration event at time $t_k$ and the $j$-th sample lineage was involved. 
  Whether this migration event was presented as a deme-transition event in $\Och_q$ depends on whether $j''=j'$.
  The population size of Deme $j$ right before time $t_k$ was $I_{k-1}^j$, while that of Deme $j'$ right after $t_k$ was $I_k^{j'}$.
  Lineages of $\Och_{q-1}$ were not involved in this migration event, so they were excluded.
  Thus, at time $t_k$, one of the $I_{k-1}^{j}-\ell_{q,k-1}^{j}$ individuals was chosen to migrate and right after, among $I_{k}^{j'}-\ell_{qk}^{j'}$ unobserved individuals, the $j$-th sample was the exact one just migrated.
  Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
  \item If $t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'}$, the logic is the same as that in case \ref{item:M2Z}, but the migration event was not presented in $\Och_q$. One can avoid the exact genealogical deme-transition induced by a migration at time $t_k$, with $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
  \item \label{item:G2D} If $t_k=a_q\in\Lset^{j}(\Och_{q-1})$, and obviously, $u_k\in\Sset_j$.
    A individual in Deme $j$ was sampled at $t_k$ and the $q$-th lineage sample was exactly the direct-descend of it, \ie a direct-descend event caused by the $q$-th lineage sample was presented in $\Och_q$ at time $t_k$.

  \item
  \item If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Bset^{j,j'}$, then there was a birth event at time $t_k$ and the $j$-th sample lineage was involved. 
    However, this birth event was observed as a deme-transition event (\ie $j\neq j'$) in $\Och_q$, instead of a coalescent event. 
    This birth event resulted in a newly added node holding two black balls, which were separately from Deme $j$ and Deme $j'$.
    Thus, the population size of Deme $j$ and $j'$, right after $t_k$, were $I_k^j$ and $I_k^{j'}$, respectively. 
    There were in total $(I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})$ combinations of the black balls, because none of the lineages in $\Och_{q-1}$ was involved. 
    Among all these combinations, the $j$-th sample lineage was exactly the black ball newly added to Deme $j'$ after the birth event.
    Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})\right)$.
  \item
  \item
\end{enumerate}

@@

    % deme-transition time can evolve to coalescent time, when the new sample is the parent of a "cross-deme" birth
    1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
    1, &\text{if}\ j''\notin\{j,j'\}\\[12pt]
    \frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
    1-\frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
    \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1}),\\[12pt]
    1-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
    \frac{\delta_{j',j''}}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]

    % B2C: either j''=j' or j''=j, the numerator = 1 when j'!=j and =2 when j'=j.
    \frac{1+\delta_{j,j'}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
    1-\frac{(1+\delta_{j,j'})\ell_{qk}^{j}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{1}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
    0, &\text{otherwise}.
  \end{cases}
\end{equation}

>>>>

All cases can be categorized into four groups based on the occurrence of population event within time $\halfopen{a_q,s_q}$: no population event or simultaneous events, a birth from Deme $j$ to $j'$, a sample of Deme $j'$, and a migration from Deme $j$ to $j'$ where $j\neq j'$. Within each group, we investigate $\phi_{qk}^{j''}$ by considering whether the corresponding population event was presented in the pruned genealogy and involving the $q$-th lineage sample at time $t_k$, given that the $q$-th lineage sample was in Deme $j''$ at time $t_k$. For $q>1$, we then have:

\begin{enumerate}[(a)]
  \item If $t_k\notin\halfopen{a_q,s_q}$, or if $u_k\notin\Bset\cup\Sset\cup\Mset$, so we have $\phi_{qk}^{j''}=1$.
  \item If $t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1})$, then again, $\phi_{qk}^{j''}=1$.
  \item \label{item:M2Z} If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Mset^{j,j'}$, then there was a migration event at time $t_k$ and the $q$-th sample lineage was involved. 
  Whether this migration event was presented as a deme-transition event in $\Och_q$ depends on whether $j''=j'$.
  The population size of Deme $j$ right before time $t_k$ was $I_{k-1}^j$, while that of Deme $j'$ right after $t_k$ was $I_k^{j'}$.
  Lineages of $\Och_{q-1}$ were not involved in this migration event, so they were excluded.
  Thus, at time $t_k$, one of the $I_{k-1}^{j}-\ell_{q,k-1}^{j}$ individuals was chosen to migrate and right after, among $I_{k}^{j'}-\ell_{qk}^{j'}$ unobserved individuals, the $q$-th sample was the exact one just migrated.
  Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
  \item If $t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'}$, the logic is the same as that in case \ref{item:M2Z}, but the migration event was not presented in $\Och_q$. One can avoid the exact genealogical deme-transition induced by a migration at time $t_k$, with $\phi_{qk}^{j''}=1-\delta_{j',j''}/\left((I_{k-1}^{j}-\ell_{q,k-1}^{j})(I_{k}^{j'}-\ell_{qk}^{j'})\right)$.
  \item \label{item:G2D} If $t_k=a_q\in\Lset^{j}(\Och_{q-1})$, and obviously, $u_k\in\Sset_j$.
    A individual in Deme $j$ was sampled at $t_k$ and the $q$-th lineage sample was exactly the direct-descend of it, \ie a direct-descend event caused by the $q$-th lineage sample was presented in $\Och_q$ at time $t_k$.

  \item
  \item If $t_k\in(a_q,s_q)\setminus\Masterset(\Och_{q-1})$, $t_k\in\Zset^{j,j'}(\Och_q)$, and $u_k\in\Bset^{j,j'}$, then there was a birth event at time $t_k$ and the $q$-th sample lineage was involved. 
    However, this birth event was observed as a deme-transition event (\ie $j\neq j'$) in $\Och_q$, instead of a coalescent event. 
    This birth event resulted in a newly added node holding two black balls, which were separately from Deme $j$ and Deme $j'$.
    Thus, the population size of Deme $j$ and $j'$, right after $t_k$, were $I_k^j$ and $I_k^{j'}$, respectively. 
    There were in total $(I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})$ combinations of the black balls, because none of the lineages in $\Och_{q-1}$ was involved. 
    Among all these combinations, the $q$-th sample lineage was exactly the black ball newly added to Deme $j'$ after the birth event.
    Therefore, $\phi_{qk}^{j''}=\delta_{j',j''}/\left((I_k^{j}-\ell_{qk}^{j})(I_k^{j'}-\ell_{qk}^{j'})\right)$.
  \item
  \item
\end{enumerate}

@@

    % deme-transition time can evolve to coalescent time, when the new sample is the parent of a "cross-deme" birth
    1, &\text{if}\ t_k\in\Cset(\Och_{q-1})\cup\Qset(\Och_{q-1}),\\[12pt]
    \frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
    1-\frac{\delta_{j',j''}}{(I_{k-1}^j-\ell_{q,k-1}^{j})(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Mset^{j,j'},\\[12pt]
    \frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k=a_q\in\Lset^{j'}(\Och_{q-1}),\\[12pt]
    1-\frac{\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q)\setminus\Qset(\Och_{q-1})\ \text{and}\ u_k\in\Sset^{j'},\\[12pt]
    \frac{\delta_{j',j''}}{(I_k^j-\ell_{qk}^j)(I_k^{j'}-\ell_{qk}^{j'})}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_q)\setminus\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]

    % B2C: either j''=j (in Z_{q-1} or j''=j' (not in Z_{q-1})
    % the numerator = 1 when j'!=j and =2 when j'=j.
    \frac{\delta_{j,j''}+\delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}, &\text{if}\ t_k=a_q\in\Cset^{j,j'}(\Och_q),\\[12pt]
    1-\frac{(\delta_{j,j''}+\delta_{j',j''})\ \ell_{qk}^{j}\ \delta_{j',j''}}{I_k^{j}(I_k^{j'}-\delta_{j,j'})-\ell_{qk}^j(\ell_{qk}^{j'}-\delta_{j,j'})}-\frac{(1-\delta_{j,j'})\delta_{j',j''}}{I_k^{j'}-\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\notin\Zset^{j,j'}(\Och_{q-1})\cup\Zset^{j,j'}(\Och_q),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
    1-\frac{\delta_{j,j''}}{I_k^{j}I_k^{j'}-\ell_{qk}^j\ell_{qk}^{j'}}, &\text{if}\ t_k\in(a_q,s_q),\ t_k\in\Zset^{j,j'}(\Och_{q-1}),\ \text{and}\ u_k\in\Bset^{j,j'},\\[12pt]
    0, &\text{otherwise}.
  \end{cases}
\end{equation}
